Bootstrapping Domain Specific Languages for Neurally-Guided Bayesian Program Learning

### 2018-02-09 paper.tex
another symbolic regression experiment curve
fixed our use of system at end of sentences

\newcommand{\system}{\textsc{DreamCoder}~}
\newcommand{\lowerBound}{\mathscr{L}}
\newcommand{\code}[1]{{\footnotesize\texttt{#1}}}
% Use the following line for the initial blind version submitted for review:
\usepackage[nohyperref]{icml2018}

@@

How should we perform inference for the model
in~\ref{graphicalModel}(a)?
We take inspiration from the Helmholtz machine and the Wake/Sleep algorithm~\cite{hinton1995wake},
which alternates between ``waking'', which draws samples from a recognition model (our $q$);
and  ``sleeping'', which trains a recognition model on samples from the generative model (our $\mathcal{D},\theta$).
For us, a wake cycle means using the recognition model to search for programs that solve the tasks.
The wake cycle can be thought of active problem solving,
guided by the current generative and recognition models.
A sleep cycle means  updating these models so that the DSL and  recognition model
more closely match the distribution of tasks.
An important difference from the Helmholtz machine
is that our setup has two distinct sleep phases: a phase where we update the generative model
$(\mathcal{D},\theta)$, which we call \textbf{Sleep-G},
and a second phase where we update the recognition model $q$,
which we call \textbf{Sleep-R}.
This wake/sleep style of inference is the reason why we call our approach \system.

@@

    \draw [->,thick,dashed] (DSL.north) to[out = 90,in = 180] node[fill=white,align = center]{  \footnotesize{Inductive bias}\\\footnotesize{(Wake)}} (synthesis.west);
    \draw [->,thick,dashed] (recognitionModel.north) to[out = 90,in = 0] node[fill=white,align = center]{{\footnotesize Makes tractable}\\{\footnotesize (Wake)}} (synthesis.east);
  \end{scope}
    \end{tikzpicture}
\caption{Our algorithm solves for programs, the DSL, and a recognition model. Each of these steps bootstrap off of the others in a Helmholtz-machine inspired wake/sleep inference algorithm.}  \label{feeding}
\end{figure}

@@

The neural architecture of $q$ depends upon the domain.
For domains with sequential structure like strings and lists,
we use an RNN to encode each input/output example,
and for domains with a fixed-dimensionality inputs/outputs we use a multilayer perceptron.
We MaxPool across all of the input/output examples. See supplement for details.

@@

\section{Experiments}
We first present our experimental results qualitatively,
illustrating the DSLs that  \system
learns in the domains of Boolean circuits, symbolic regression, string editing, and list functions.
We then show quantitative results (Sec.~\ref{quantitative})
comparing \system with lesioned versions
of our model that are analogous to prior work.

@@

What DSL does \system learn?
The learned DSL contains templates for polynomials of different orders,
which lets the algorithm quickly hone in on the kinds of functions that are most appropriate to this domain (Tbl.~\ref{regressionDSL}).
Examining the programs themselves,
one finds that the algorithm discovers representations for each of the polynomials that minimizes the number of continuous degrees of freedom:
for example, it represents the linear function $-x+2$ with the program 
\code{(* (* (+ x }$\mathcal{R}$\code{) }$\mathcal{R}$\code{) x)}, which has two continuous degrees of freedom -- while representing quartic functions using the invented DSL primitive $f_6$ in Tbl.~\ref{regressionDSL}.
This falls out of our Bayesian framing -- both the bias towards shorter programs (more likely under $\mathcal{D},\theta$) and the likelihood model's BIC penalty.
\begin{table}\centering
\begin{tabular}{l}
  \toprule
  %% Primitives& $+,\times:\mathbb{R}\to \mathbb{R}\to \mathbb{R}$\\
  %% & $\mathcal{R}:\mathbb{R}$ (real valued parameter)\\\midrule
  %% Observation $x$& $N$ input/output examples: $\left\{(i_n,o_n) \right\}_{n\leq N}$\\
  %% \midrule
  %% Likelihood $\probability[x|p]$& $\propto\exp\left( - D\log N \right)\prod_{n\leq N}\mathcal{N}(p(i_n)|o_n)$
  %% \\\midrule
    $f_0 = $\code{(+ }$\mathcal{R}$\code{)}\\
    $f_1($\code{x}$) =$ \code{(* (* (+ x }$\mathcal{R}$\code{) }$\mathcal{R}$\code{) x)} \\
      $f_2($\code{x}$) =$ \code{(+ }$\mathcal{R}$\code{ ($f_1$ x))} \\
      $f_3($\code{x}$) = $\code{(* x (}$f_2$\code{ x))}\\
  $f_4($\code{x}$) = $\code{(}$f_0$\code{ (}$f_3$\code{ x)))}\\
  $f_5($\code{x}$) =$ \code{(* x (}$f_4$\code{ x))}\\
  $f_6($\code{x}$) = $\code{(}$f_0$\code{ (}$f_4$\code{ x))} (\emph{4th order polynomial})
  \\\bottomrule
\end{tabular}
\caption{Some induced DSL primitives for symbolic regression. System starts out with addition, multiplication, and real numbers, and learns functions for building 4th order polynomials.}\label{regressionDSL}
\end{table}

@@

\subsection{String editing}\label{textSection}
Synthesizing programs that manipulate strings is a classic problem in the
programming languages and AI literatures~\cite{menon2013machine,lau2001programming},
and algorithms that learn string editing programs ship in Microsoft Excel~\cite{gulwani2011automating}.
However, this prior work presumes a ready-made DSL,
expertly crafted to suit string editing.
We show in this experiment that we can instead start out with a
less specialized set of primitives -- close to the string routines provided by Python --
and recover many of the higher-level building blocks that have made these
other system successful.
Tbl.~\ref{textPrimitives} (top) lists the primitive functions provided to \system.
Absent are any routines for searching for characters, extracting substrings,
or replacing one character with another.
Instead, we provide list/string indexing primitives (\code{slice}, \code{nth}),
basic arithmetic, and functions for splitting a string into a list (\code{split})
or turning a list into a string (\code{join}).
We automatically generated  600 string editing tasks along with input/output examples (Fig.~\ref{exampleTextProblem}). %, split  75/25 train/test.
At first, \system cannot find any programs that
work at all for most of the tasks.
After three wake/sleep iterations,
it assembles a DSL (Tbl.~\ref{textPrimitives}, bottom) that lets it rapidly explore the space of programs and find solutions to
all of the tasks.

\begin{table}[t]
  \centering
  \begin{tabular}{c}
    String editing primitives\\
    \begin{tabular}{l}
    \toprule
    \code{0}, \code{+1},\code{-1},     \code{""},    \code{","},     \code{" "},    \code{"<"}, \code{">"}  \\
    \code{split}, \code{join}, \code{map}, 
    \code{append}, \code{chr->str}, \code{trim}, \code{nth}\\
    \code{length}, \code{slice},    \code{lower}, \code{upper}, \code{capitalize}
  \\\bottomrule
    \end{tabular}\\
\\    Subset of learned DSL\\
    \begin{tabular}{l}
      \toprule
      $f_0($\code{c}$) = $\code{(+ (chr->str c))}\\
      $f_1($\code{i,c,s}$) = $      \code{(+ (nth i (split c s))}\\\hspace{2.55cm}\code{ (chr->str c))}\\
      $f_2($\code{s,c}$) = $\code{((}$f_0$\code{ c) (}$f_1$\code{ (1+ 0) c s))} \\
      \hspace{0.5cm}($f_2$: \emph{Extract substring delimited by character})\\
      $f_3($\code{c,s}$) = $\code{(nth (1+ 0) (split c (nth 0 s)))}\\
      \hspace{0.5cm}($f_3$: \emph{2nd substring delimited by character})
    %% $f_0($\code{i,c,s}$) = $    \code{(+ (index i (split c s)) (chr->str c))}\\
    %% $f_1 = $\code{(join "")}
  \\\bottomrule
    \end{tabular}
  \end{tabular}
  \caption{Top: provided string editing primitives. Bottom: some induced DSL primitives.}\label{textPrimitives}
  \end{table}

@@

  \\\midrule\multicolumn{6}{c}{\emph{Symbolic Regression}}\\\midrule
  \% solved&   \textbf{ 98\% }&94\%&0.7\%&0\%&2\% \\
  Solve time&  2.7s& 2.8s  &3.6s&--&2.1s\\
  MDL (nats)&   9.3 &11.0&2.3&--&4.6

@@

  \\\midrule\multicolumn{6}{c}{\emph{List functions (rich DSL)}}\\\midrule
  \% solved&\textbf{86\%} &84\% &60\%&74\%&70\%\\
  Solve time&  0.8s&0.7s&1.0s&1.0s&1.1s
  \\\bottomrule
  \end{tabular}
\caption{\% solved: fraction solved with 5 sec timeout. Solve time: averaged over solved tasks. MDL: $-\expect\left[\probability[x|p] \right]$. For domains other than symbolic regression MDL is 0 nats. For symbolic regression MDL is a proxy for \# of continuous degrees of freedom.}\label{baselineComparisons}  \end{table}

@@

 \section{Conclusion}
 We contributed a model, \system, that learns to program by
 bootstrapping a DSL with new domain-specific primitives that the algorithm itself discovers.
  Alongside this DSL, we train a bottom-up
 neural recognition model that learns how to efficiently deploy the
 DSL and hone in on the right programs.  We believe that this
 integration of top-down symbolic representation and bottom-up neural
 networks -- both of them learned -- is a useful general approach for
 getting program induction off the ground.

>>>>

\newcommand{\system}{\textsc{DreamCoder}~}
\newcommand{\endingSystem}{\textsc{DreamCoder}}
\newcommand{\lowerBound}{\mathscr{L}}
\newcommand{\code}[1]{{\footnotesize\texttt{#1}}}
% Use the following line for the initial blind version submitted for review:
\usepackage[nohyperref]{icml2018}

@@

How should we perform inference for the model
in~\ref{graphicalModel}(a)?
We take inspiration from the Helmholtz machine and the Wake/Sleep algorithm~\cite{hinton1995wake},
which alternates between ``waking'', which draws samples from a recognition model (our $q$);
and  ``sleeping'', which trains a recognition model on samples from the generative model (our $\mathcal{D},\theta$).
For us, a wake cycle means using the recognition model to search for programs that solve the tasks.
The wake cycle can be thought of active problem solving,
guided by the current generative and recognition models.
A sleep cycle means  updating these models so that the DSL and  recognition model
more closely match the distribution of tasks.
An important difference from the Helmholtz machine
is that our setup has two distinct sleep phases: a phase where we update the generative model
$(\mathcal{D},\theta)$, which we call \textbf{Sleep-G},
and a second phase where we update the recognition model $q$,
which we call \textbf{Sleep-R}.
This wake/sleep style of inference is the reason why we call our approach \endingSystem.

@@

    \draw [->,thick,dashed] (DSL.north) to[out = 90,in = 180] node[fill=white,align = center]{  \footnotesize{Inductive bias}\\\footnotesize{(Wake)}} (synthesis.west);
    \draw [->,thick,dashed] (recognitionModel.north) to[out = 90,in = 0] node[fill=white,align = center]{{\footnotesize Makes tractable}\\{\footnotesize (Wake)}} (synthesis.east);
  \end{scope}
    \end{tikzpicture}
  \caption{\system solves for programs, the DSL, and a recognition model. Each of these steps bootstrap off of the others in a Helmholtz-machine inspired wake/sleep inference algorithm.}  \label{feeding}
\end{figure}

@@

The architecture of $q$ depends upon the domain.
For domains with sequential structure like strings and lists,
we use an RNN to encode each input/output example,
and for domains with fixed-dimensionality inputs/outputs we use a multilayer perceptron.
We MaxPool across all of the input/output examples. See supplement for details.

@@

\section{Experiments}
We first present our experimental results qualitatively,
illustrating the DSLs we
learn.
We then show quantitative results (Sec.~\ref{quantitative})
comparing \system with lesioned versions
of our model that are analogous to prior work.

@@

What DSL does \system learn?
The learned DSL contains templates for polynomials of different orders,
which lets the algorithm quickly hone in on the kinds of functions that are most appropriate to this domain (Tbl.~\ref{regressionDSL}).
Examining the programs themselves,
one finds that the algorithm discovers representations for each of the polynomials that minimizes the number of continuous degrees of freedom:
for example, it represents the linear function $-x+2$ with the program 
\code{(* (* (+ x }$\mathcal{R}$\code{) }$\mathcal{R}$\code{) x)}, which has two continuous degrees of freedom -- while representing quartic functions using the invented DSL primitive $f_6$ in Tbl.~\ref{regressionDSL}.
This falls out of our Bayesian framing -- both the bias towards shorter programs (more likely under $\mathcal{D},\theta$) and the likelihood model's BIC penalty.
\begin{table}\centering
\begin{tabular}{l}
  \toprule
  %% Primitives& $+,\times:\mathbb{R}\to \mathbb{R}\to \mathbb{R}$\\
  %% & $\mathcal{R}:\mathbb{R}$ (real valued parameter)\\\midrule
  %% Observation $x$& $N$ input/output examples: $\left\{(i_n,o_n) \right\}_{n\leq N}$\\
  %% \midrule
  %% Likelihood $\probability[x|p]$& $\propto\exp\left( - D\log N \right)\prod_{n\leq N}\mathcal{N}(p(i_n)|o_n)$
  %% \\\midrule
    $f_0 = $\code{(+ }$\mathcal{R}$\code{)}\\
    $f_1($\code{x}$) =$ \code{(* (* (+ x }$\mathcal{R}$\code{) }$\mathcal{R}$\code{) x)} \\
      $f_2($\code{x}$) =$ \code{(+ }$\mathcal{R}$\code{ ($f_1$ x))} \\
      $f_3($\code{x}$) = $\code{(* x (}$f_2$\code{ x))}\\
  $f_4($\code{x}$) = $\code{(}$f_0$\code{ (}$f_3$\code{ x)))}\\
  $f_5($\code{x}$) =$ \code{(* x (}$f_4$\code{ x))}\\
  $f_6($\code{x}$) = $\code{(}$f_0$\code{ (}$f_4$\code{ x))} (\emph{4th order polynomial})
  \\\bottomrule
\end{tabular}
\caption{Some induced DSL primitives for symbolic regression. System starts out with addition, multiplication, and real numbers, and learns to build 4th order polynomials.}\label{regressionDSL}
\end{table}

@@

\subsection{String editing}\label{textSection}
Synthesizing programs that manipulate strings is a classic problem in the
programming languages and AI literatures~\cite{menon2013machine,lau2001programming},
and algorithms that learn string editing programs ship in Microsoft Excel~\cite{gulwani2011automating}.
However, this prior work presumes a ready-made DSL,
expertly crafted to suit string editing.
We show in this experiment that we can instead start out with a
less specialized set of primitives -- close to the string routines provided by Python --
and recover many of the higher-level building blocks that have made these
other system successful.
Tbl.~\ref{textPrimitives} (top) lists the primitive functions provided to \endingSystem.
Absent are any routines for searching for characters, extracting substrings,
or replacing one character with another.
Instead, we provide list/string indexing primitives (\code{slice}, \code{nth}),
basic arithmetic, and functions for splitting a string into a list (\code{split})
or turning a list into a string (\code{join}).
We automatically generated  600 string editing tasks along with input/output examples (Fig.~\ref{exampleTextProblem}). %, split  75/25 train/test.
At first, \system cannot find any programs that
work at all for most of the tasks.
After three wake/sleep iterations,
it assembles a DSL (Tbl.~\ref{textPrimitives}, bottom) that lets it rapidly explore the space of programs and find solutions to
all of the tasks.

\begin{table}[t]
  \centering
  \begin{tabular}{c}
    String editing primitives\\
    \begin{tabular}{l}
    \toprule
    \code{0}, \code{+1},\code{-1},     \code{""},    \code{","},     \code{" "},    \code{"<"}, \code{">"}  \\
    \code{split}, \code{join}, \code{map}, 
    \code{+}, \code{chr->str}, \code{trim}, \code{nth}\\
    \code{length}, \code{slice},    \code{lower}, \code{upper}, \code{capitalize}
  \\\bottomrule
    \end{tabular}\\
\\    Subset of learned DSL\\
    \begin{tabular}{l}
      \toprule
      $f_0($\code{c}$) = $\code{(+ (chr->str c))}\\
      $f_1($\code{i,c,s}$) = $      \code{(+ (nth i (split c s))}\\\hspace{2.55cm}\code{ (chr->str c))}\\
      $f_2($\code{s,c}$) = $\code{((}$f_0$\code{ c) (}$f_1$\code{ (1+ 0) c s))} \\
      \hspace{0.5cm}($f_2$: \emph{Extract substring delimited by character})\\
      $f_3($\code{c,s}$) = $\code{(nth (1+ 0) (split c (nth 0 s)))}\\
      \hspace{0.5cm}($f_3$: \emph{2nd substring delimited by character})
    %% $f_0($\code{i,c,s}$) = $    \code{(+ (index i (split c s)) (chr->str c))}\\
    %% $f_1 = $\code{(join "")}
  \\\bottomrule
    \end{tabular}
  \end{tabular}
  \caption{Top: provided string editing primitives. Bottom: some induced DSL primitives.}\label{textPrimitives}
  \end{table}

@@

  \\\midrule\multicolumn{6}{c}{\emph{Symbolic Regression}}\\\midrule
  \% solved&   \textbf{98\% }&94\%&0.7\%&0\%&2\% \\
  Solve time&  2.7s& 2.8s  &3.6s&--&2.1s\\
  MDL (nats)&   9.3 &11.0&2.3&--&4.6

@@

  \\\midrule\multicolumn{6}{c}{\emph{List functions (rich DSL)}}\\\midrule
  \% solved&\textbf{86\%} &84\% &60\%&74\%&70\%\\
  Solve time&  0.8s&0.7s&1.0s&1.0s&1.1s
  \\\bottomrule
  \end{tabular}
\caption{\% solved: fraction solved w/ 5 sec timeout. Solve time: averaged over solved tasks. MDL: $-\expect\left[\probability[x|p] \right]$. For domains other than symbolic regression MDL is 0 nats. For symbolic regression MDL is a proxy for \# of continuous degrees of freedom.}\label{baselineComparisons}  \end{table}

@@

 \section{Contribution and Outlook}
 We contributed a model, \system, that learns to program by
 bootstrapping a DSL with new domain-specific primitives that the algorithm itself discovers.
  Alongside this DSL, we train a bottom-up
 neural recognition model that learns how to efficiently deploy the
 DSL and hone in on the right programs.  We believe that this
 integration of top-down symbolic representation and bottom-up neural
 networks -- both of them learned -- is a useful general approach for
 getting program induction off the ground.
